Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-27 article.tex
forgot reference to other Hutton parsing paper
Responses.
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

\EZY{I rewrote the abstract, let me know if you like it.}
\Matt{LGTM, thanks!}

@@

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}
Note that the FlexibleContexts language extension is required by the 
MonadState constraint in the type signature.  The mtl \EZY{which?} \Matt{I'm not sure what the library's called -- is it mtl or transformers?} formulation of monad
transformers relies heavily on extensions to the type system, more of which 
we'll have to use for some of the later code in this article.

@@

Of course not -- we haven't told it how to skip comments and whitespace yet!
Let's do that now, first by defining the whitespace and comment patterns, then
by creating a \verb+munch+ combinator that takes in a parser, 
throws away all leading junk, and finally runs the parser.  Also, we rename
the \verb+some+ and \verb+many+ functions from Control.Alternative to more
clearly indicate their semantics: \EZY{I would suspect many compiler authors would be of the opinion that the proper way to remove comments and whitespace is to lex first, and parse later.}
\Matt{That's a good point, I toyed with doing it that way but it was just too much trouble for no benefit that I could see.  It ended up being a lot more code.
But this approach is equivalent, since the token parsers get the first crack at the input string, and the non-token parsers don't touch the input string directly.}
\begin{verbatim}
many0 = many
many1 = some

@@

This combinator takes two arguments: an error value and a monadic 
computation.  It first tries to run the computation; if the computation 
fails, it generates an error.  The 
definition of 'failure' is provided by the Alternative instance of \verb+m+.
\EZY{Give a little intuition on why the operator is called commit earlier;
you don't get around it until later.}
The reason for its name is that it restricts backtracking and alternatives;
it ``commits" the parser to a do-or-die situation -- either it successfully
parses the input, or the parsing stops.

@@

On the other hand, the order of Error and Maybe relative to each other isn't 
important.  But don't just take my word for it -- try it for yourself!
\EZY{It's worth mentioning why, formally speaking, this is the case, which has
to do with the relevant structures being isomorphic to each other.}
\Matt{Uh-oh!  I don't really know anything formal about monad transformers,
I really have no idea how to go about showing something like this!
I guess I'll have to rethink my goals for this section.}

>>>>

@@

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}
Note that the FlexibleContexts language extension is required by the 
MonadState constraint in the type signature.  The mtl \EZY{which?} \Matt{I'm not sure what the library's called -- is it mtl or transformers?} \EZY{Well, both export the necessary interfaces. Just pick one and be consistent.} formulation of monad
transformers relies heavily on extensions to the type system, more of which 
we'll have to use for some of the later code in this article.

@@

Of course not -- we haven't told it how to skip comments and whitespace yet!
Let's do that now, first by defining the whitespace and comment patterns, then
by creating a \verb+munch+ combinator that takes in a parser, 
throws away all leading junk, and finally runs the parser.  Also, we rename
the \verb+some+ and \verb+many+ functions from Control.Alternative to more
clearly indicate their semantics:
\begin{verbatim}
many0 = many
many1 = some

@@

This combinator takes two arguments: an error value and a monadic 
computation.  It first tries to run the computation; if the computation 
fails, it generates an error.  The 
definition of 'failure' is provided by the Alternative instance of \verb+m+.
\EZY{Give a little intuition on why the operator is called commit earlier;
you don't get around it until later.}
The reason for its name is that it restricts backtracking and alternatives;
it ``commits" the parser to a do-or-die situation -- either it successfully
parses the input, or the parsing stops.  It's based on the \verb+nofail+
combinator of \cite{hutton1992higher}.

@@

On the other hand, the order of Error and Maybe relative to each other isn't 
important.  But don't just take my word for it -- try it for yourself!
\EZY{It's worth mentioning why, formally speaking, this is the case, which has
to do with the relevant structures being isomorphic to each other.}
\Matt{Uh-oh!  I don't really know anything formal about monad transformers,
I really have no idea how to go about showing something like this!
I guess I'll have to rethink my goals for this section.}
\EZY{It's pretty simple, and just has to do with the fact that Maybe (Either a b) is
equivalent (modulo laziness) with Either a (Maybe b).  To show this is a case, you have
to define functions which convert from one to the other, and then show that the roundtrip
is identity.}
